---
title: "Exam 1"
editor: visual
---
```{r}
library(tidytuesdayR)
library(tidyverse)
```


```{r}
# Get the Data



tuesdata <- tidytuesdayR::tt_load('2020-02-18')
tuesdata <- tidytuesdayR::tt_load(2020, week = 8)


food_consumption <- tuesdata$food_consumption
```
```{r}
dim(food_consumption)
#nrow(food_consumption)
#ncol(food_consumption)
head(food_consumption)
names(food_consumption)
structure(food_consumption)
tail(food_consumption)
```
```{r}
#Shows first 22 rows
head(food_consumption, 22)
```
```{r}
#Shows last 22 rows
tail(food_consumption, 22)
```

There are 4 variables
 "country"
 "food_category"
 "consumption"
 "co2_emmission"
 
 What does consumption of each food category in each country look like?
 
A spacial plot wiht a map showing the size if points with the amount of condumption and the color on the points representing the food category. 
 
geom_tile = table with colors

Concepts to make final viz more effective

1.) labels
2.) alternate text (fig-alt)
3.) color blind friendliness (viridis_d())
4.) text captions (fig-cap)
5.) controlling height of plot (fig-height)
6.) controlling width (fig-width)

```{r}
library(ggplot2)
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

world <- ne_countries(scale = "medium", returnclass = "sf") 

world_coords <- world %>% 
  select(name, geometry) %>% 
  mutate(centroid = st_centroid(geometry),
         lon = st_coordinates(centroid)[,1],
         lat = st_coordinates(centroid)[,2]) %>%
  select(name, lon, lat)

food_map_data <- food_consumption %>%
  left_join(world_coords, by = c("country" = "name")) %>%
  filter(!is.na(lon) & !is.na(lat))  

ggplot() +
  geom_sf(data = world, fill = "gray90", color = "white") +  # Base map
  geom_point(data = food_map_data, 
             aes(x = lon, y = lat, size = consumption, color = food_category),
             alpha = 0.7) +  
  scale_size(range = c(1, 10)) +  
  labs(title = "Global Food Consumption", subtitle = "(Size = Consumption, Color = Food Category)", x = "Longitude", y = "Latitude") +
  theme_minimal()
  
```
My study sheet images are below. 
(I couldnt get them to upload in but I do have them done and copies saved)

```{r}
#knitr::include_graphics("STAT112 (exam 1) study sheet pt.1.png")
#knitr::include_graphics("STAT112  (exam 1) study sheet pt.2.png")
#knitr::include_graphics("STAT112  (exam 1) study sheet pt.3.png")


```


